%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------


\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% numbers       To obtain numeric citation style instead of author/year.

\usepackage{amsmath}
\usepackage{haskell}
\usepackage{todo}
\usepackage{multicol}
\usepackage{multirow}


\newcommand{\cL}{{\cal L}}

\def \meet {\<meet\> ($\sqcap$) }
%\def \pmeet {\bm{\&}}
\def \join {\<join\> ($\sqcup$) }
\DeclareMathOperator{\pmeet}{\&}


\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\copyrightdoi{nnnnnnn.nnnnnnn}

% Uncomment the publication rights you want to use.
%\publicationrights{transferred}
%\publicationrights{licensed}     % this is the default
%\publicationrights{author-pays}

\titlebanner{DRAFT}        % These are ignored unless
\preprintfooter{DRAFT}   % 'preprint' option specified.

\title{Demanding Sums}
\subtitle{Extending Higher-Order Demand Analysis to Sum Types}

\authorinfo{Name1}
           {Affiliation1}
           {Email1}
\authorinfo{Name2\and Name3}
           {Affiliation2/3}
           {Email2/3}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore,
% you may leave them out
\terms
term1, term2

\keywords
keyword1, keyword2

\section{Introduction}

There are two main motivations for this work: being able to derive parallel
strategies for higher-order programs (JMCT's previous work only derived
strategies for first-order programs), and the ability to perform a
worker/wrapper like optimisation on sum types. Both of these goals have one
major obstacle in their way, the lack of a higher-order demand analysis that
works on sum types. This paper aims to introduce such an analysis, and show how
it can be used for these two goals.

The success of the worker/wrapper transform has been one of the driving factors
in the use and improvement of strictness analysis in lazy functional languages.
Because of this fact GHC, which has a very sophisticated stricntess analyser
(called demand analysis in GHC) focuses on product types and does not
discriminate between the constructors of a sum type. This is fine for the
work/wrapper transfor as it has been applied in the past. Part of the reason
that sum types were not a focus of past analysis efforts is because it is not
immediately clear how to unbox a sum type. Luckily Johan Tibell has worked on a
method for unboxing sum types, removing unnecessary indirections in the heap.
Tibell's approach is \emph{manual}, the programmer must specify which types are
needing this representation. By extending the demand analysis to sum types, we
would be able to provide the compiler with enough information to make this
determination for us. much like it does for product types in the worker/wrapper
transformation.

One of the recent attempts at implicit parallelism for lazy languages used a
demand analysis on first-oder programs to \emph{derive} parallel strategies.
The technique relies on the analyses ability to provide differing demands on
different constructors from a sum type.  This way the derived strategy can
perform case analysis on sum types and evaluate to different depths with the
differing demands.

\subsection*{Roadmap}

In 

\section{Demand Analysis}
\label{sec:demandReview}
What is meant by `demand'? As an example, the function \verb'length' requires
that the input list be finite, but no more. We can therefore say that
\verb'length' \emph{demands} the spine of the argument list. The function
\verb'append' is a more interesting example:

\begin{haskell*}
append &::& [\alpha] \to [\alpha] \to [\alpha] \\
append []     ys &=& ys \\
append (x:xs) ys &=& x : append xs ys
\end{haskell*}

As mentioned in the previous section the first argument must be defined to the
first cons, but we cannot know whether the second argument is ever needed.
However, what if the \emph{result} of \verb'append' needs to be a finite list?
For example:

\begin{haskell*}
lengthOfBoth &::& [\alpha] \to [\alpha] \to Int \\
lengthOfBoth xs ys &=& length (append xs ys)
\end{haskell*}

In this case \emph{both} arguments to \verb'append' must be finite. Projections
can be used to formalise this type of context \citep{wadler1987projections,
hinze1995projection}, which we call a \emph{demand context}.

Demand Contexts allow us to reason about the \emph{use} of a function's result,
letting us reason about functions like \<append\> more accurately. This, combined
with their ability to analyse functions of arbitrary types without the need to
design abstract domains by hand, make projection-based analysis the most
realistic for our purposes of utilising implicit parallelism.

\subsection{Semantics of Projections}
\label{sec:projSem}

Given a domain $D$, a projection on $D$ is a continuous function
$\pi \ : \ D \rightarrow D$ that satisfies

\begin{align}
\pi \sqsubseteq ID \\
\pi \circ \pi = \pi
\end{align}

Equation (3) ensures that a projection can not add any information to a value,
i.e. all projections approximate the identity function. Idempotence (4) ensures
that projecting the same demand twice on a value has no additional effect. This
aligns with our intuition of demand. If we demand that a list is spine-strict,
demanding spine-strictness again does not change the demand on the list.

Because we want the introduction of parallelism to be semantics-preserving we
use the following safety condition for projections:

\begin{equation}
\gamma \ \circ \ f = \gamma \ \circ \ f \ \circ \ \pi
\end{equation}

Given a function $f \ : X \rightarrow Y$, and demand $\gamma$ on the
\emph{result} of $f$, projection-based analysis propagates the demand given by
$\gamma$ to the arguments of $f$. This results in the demand on the
\emph{arguments} of $f$ given by $\pi$.  The analysis aims to find the
\emph{smallest} $\pi$ for each $\gamma$, but approximating towards $ID$ (as
it is always safe to project the identity).

\paragraph{Demands on Primitives}
On unlifted base types, such as unboxed integers, there are two demands,
$ID$ and $BOT$, with the following semantics


\begin{align}
ID \ x \ &= \ x \\
BOT \ x \ &= \ \bot
\end{align}


When an expression is in a $BOT$ context it means that non-termination is
inevitable. You can safely evaluate an expression in this context because there
is no danger of \emph{introducing} non-termination that is not already present.

\paragraph{Demands on Lifted Types} Haskell's non-strict semantics means that
most types we encounter are \emph{lifted} types.  Lifted types represent
possibly unevaluated values. Given a demand $\pi$ on $D$, we can form two
possible demands on $D_{\bot}$, $\pi!$ and $\pi?$; strict lift and lazy lift
respectively. To paraphrase Kubiak et al.: $\pi!$ means we will definitely need
the value demanded by this projection, and we will need $\pi$'s worth of it
\citep{kubiak}. $\pi?$ does not tell us whether we need the value or not, but if
we \emph{do} need the value, we will need it to satisfy $\pi$'s demand.

\paragraph{Demands on Products} A projection representing a demand on a product
can be formed by using the $\otimes$ operator with the following semantics

\begin{align*}
\langle \pi_{1} \otimes \dots \otimes \pi_{n} \rangle \ \bot &= \bot \\
\langle \pi_{1} \otimes \dots \otimes \pi_{n} \rangle \ 
\langle x_{1}, \dots, x_{n} \rangle &= \langle \pi_{1} x_{1}, \dots, \pi_{n} x_{n} \rangle
\end{align*}

\paragraph{Demands on Sums} If projections are functions on a domain, then
\nolinebreak $|$, the operator that forms projections on sum-types performs the
case-analysis. Each summand is tagged with the constructor it corresponds to.
Sometimes we will omit the constructor name when presenting projections on
types with a single summand (such as anonymous tuples).

\begin{align*}
[True\ ID | False\ ID]  \ True &= True \\
[True\ ID | False\ BOT] \ False &= \bot
\end{align*}

\begin{figure}
\begin{align*}
    \pi ::=&\ BOT              & \text{Bottom (hyperstrict)} \\
        |&\ ID               & \text{Top (the identity)} \\
        |&\ \langle \pi_{1} \otimes \pi_{2} \dots \otimes \pi_{n} \rangle   & \text{Products} \\ 
        |&\ [C_{1} \ \pi_{1} | C_{2} \ \pi_{2} \dots | C_{n} \ \pi_{n}]    & \text{Sums} \\ 
        |&\ \mu\beta . \pi     & \text{Recursive Demands} \\
        |&\ \beta              & \text{Recursive Binding} \\
        |&\ \pi?               & \text{Strict Lift} \\
        |&\ \pi!               & \text{Lazy Lift}
\end{align*}
\caption{Abstract Syntax for Contexts of Demand}
\label{fig:ContextAST}
\end{figure}


Figure \ref{fig:ContextAST} \todo{Add CVars and CRec to AST} presents a
suitable abstract syntax for projections representing demand.  This form was
introduced by Kubiak et al. and used in Hinze's work on projection-based
analyses \citep{kubiak, hinze1995projection}.  We have omitted the details on
the representation of context variables (for polymorphic demands). For a
comprehensive discussion we suggest Chapter 6 of Hinze's dissertation
\citep{hinze1995projection}.

In short, projections representing demand give us information about how defined
a value must be to satisfy a function's demand on that value. Knowing that a
value is definitely needed, and to what degree, allows us to evaluate the value
before entering the function.

\subsection*{Example Projections}

Because our primitives can be modelled by a flat domain (just $ID$ and $BOT$),
our lattice of projections corresponds with the two-point domain used in
abstract interpretation.

\hfill$\Box$

For pairs of primitive values, possible contexts include:
\begin{align}
[\langle ID? \otimes ID? \rangle] \label{IDPairs} \\
[\langle ID! \otimes ID? \rangle] \label{FSTPairs}
\end{align}


As Haskell's types are sums of products, pairs are treated as sums with only
one constructor.  For product types each member of the product is lifted.
Context \ref{IDPairs} is the top of the lattice for pairs, accepting all
possible pairs. Context \ref{FSTPairs} requires that the first member be
defined but does not require the second element. This is the demand that
\verb-fst- places on its argument.

\hfill$\Box$

For polymorphic lists there are 7 principal contexts; 3 commonly occurring contexts are:

\begin{align}
    \mu\beta.&[Nil\ ID | Cons\ \langle \gamma? \otimes \beta?\rangle] \label{IDList} \\
    \mu\beta.&[Nil\ ID | Cons\ \langle \gamma? \otimes \beta!\rangle] \label{FINList} \\
    \mu\beta.&[Nil\ ID | Cons\ \langle \gamma! \otimes \beta!\rangle] \label{FULLList}
\end{align}


Here $\mu$ binds the name for the `recursive call' of the projection and
$\gamma$ is used to represent an appropriate demand for the element type of the
list.  An important point is that this representation for recursive contexts
restricts the representable contexts to \emph{uniform projections}: projections
that define the same degree of evaluation on each of their recursive components
as they do on the structure as a whole. The detailed reason for this
restriction is given on page 89 of \cite{hinze1995projection}. This limitation
does not hinder the analysis significantly as many functions on recursive
structures are themselves uniform.

With this in mind, Context \ref{IDList} represents a lazy demand on the list,
Context \ref{FINList} represents a \emph{tail strict} demand, and Context
\ref{FULLList} represents a \emph{head and tail} strict demand on the list.

\hfill$\Box$

It will be useful to have abbreviations for a few of the contexts on lists. These
abbreviations are presented in Figure \ref{fig:listContexts}.

\begin{figure}[h!]
\begin{itemize}
    \item[] \<ID\>: accepts all lists
    \item[] \<T\> (tail strict): accepts all finite lists
    \item[] \<H\> (head strict): accepts lists where the head is defined
    \item[] \<HT\>: accepts finite lists where every member is defined
\end{itemize}
\caption[Projections for the 4-point Domain]{Four contexts on lists as described in \citep{wadler1987projections}.}
\label{fig:listContexts}
\end{figure}

We can now say more about the strictness properties of \verb'append'. The
strictness properties of a function are presented as a \emph{context
transformer} \citep{hinze1995projection}. 

\begin{align*}
    &append(ID) &\rightarrow &&ID!&;ID? \\
    &append(T)  &\rightarrow &&T!&;T! \\
    &append(H)  &\rightarrow &&H!&;H? \\
    &append(HT) &\rightarrow &&HT!&;HT!
\end{align*}

This can be read as ``If the demand on the result of \verb-append- is $ID$
then the first argument is strict with the demand $ID$ and the second
argument is lazy, but if it \emph{is} needed, it is with demand $ID$.

\hfill$\Box$

Following Hinze \citep{hinze1995projection} we construct projections
for every user-defined type. Each projection represents a
specific strategy for evaluating the structure, as we shall define in section
\ref{sec:derivation}. This provides us with the ability to generate
appropriate parallel strategies for arbitrary types. Using a
projection-based strictness analysis, we avoid the exponential blowup
of domains required for abstract interpretation.

\subsection{Lattice of Projections}

Having an intuition  of what projections are we can now define how we combine
differing demands on values. In the previous analyses we used the \meet and
\join operations directly. Projections also have \meet and \join, but because
our projections are representing demand contexts we do not actually want to use
\meet. Instead we use $\pmeet$, where $\alpha\ \pmeet\ \gamma$ represents the
\emph{joint} demand of both $\alpha$'s and $\gamma$'s demand taken together. In
other words, the projection $\alpha\ \pmeet\ \gamma$ only accepts values that
are accepted by $\alpha$ and $\gamma$, and returns $\bot$ otherwise (motivating
the use of `conjunction' to describe the operation). Using $\pmeet$ instead of
$\sqcap$ is standard when dealing with projections that act on demands
\citep{wadler1987projections, hinze1995projection, SergeyDemand}.

Figure \ref{fig:conjDisBasic} shows the rules for performing conjunction and
disjunction of projections on basic values (either $BOT$ or $ID$) and for
lifted values. Note that when we perform $\pmeet$ on two projections with
different lifts we must ensure that the resulting projection is not more strict
than the strictly lifted input, this ensures that we maintain our desired safety
condition.

\begin{figure}
\begin{multicols}{2}
\noindent

\begin{align*}
BOT       &\pmeet  \gamma        &=&\quad BOT \\
\gamma    &\pmeet  BOT           &=&\quad BOT \\
ID        &\pmeet  ID            &=&\quad ID \\
\quad &\             &\ &  \\
\alpha!        &\pmeet  \gamma!  &=&\quad (\alpha \pmeet \gamma)! \\
\alpha!        &\pmeet  \gamma?  &=&\quad (\alpha \sqcup \alpha \pmeet \gamma)! \\
\alpha?        &\pmeet  \gamma!  &=&\quad (\alpha \pmeet \gamma \sqcup \gamma)! \\
\alpha?        &\pmeet  \gamma?  &=&\quad (\alpha \sqcup \gamma)?
\end{align*}% what?

\begin{align*}
BOT            &\sqcup\  \gamma   &=&\quad \gamma \\
\gamma         &\sqcup\  BOT      &=&\quad \gamma \\
ID             &\sqcup\  ID       &=&\quad ID \\
\quad \ \quad  &                  &\ & \\
\alpha!        &\sqcup\  \gamma!  &=&\quad (\alpha \sqcup \gamma)! \\
\alpha!        &\sqcup\  \gamma?  &=&\quad (\alpha \sqcup \gamma)? \\
\alpha?        &\sqcup\  \gamma!  &=&\quad (\alpha \sqcup \gamma)? \\
\alpha?        &\sqcup\  \gamma?  &=&\quad (\alpha \sqcup \gamma)?
\end{align*}
\end{multicols}
\caption[Conjunction and Disjunction for Projections 1]{Conjunction $\pmeet$ and Disjunction $\sqcup$ for Projections on Basic and Lifted Values}
\label{fig:conjDisBasic}
\end{figure}

Figure \ref{fig:conjDisSum} shows the same operations for projections on sum
and product types. The only surprising aspect of the definitions is that we are
forced to normalise the result of a conjunction on product types. This is
because it possible for $\pmeet$ to form a projection denoting $BOT$ even when
both arguments are not $BOT$. For example, applying $\&$ to a projection that
only accepts $True$ and a projection that only accepts $False$ results in the
$BOT$ projection. This is because there is no possible Boolean value that the
resulting projection will accept, despite neither constituent projection
denoting $BOT$.

The \<norm\> function recognises these projections and transforms them to the
direct representation of $BOT$\footnote{$norm$ is defined in
\cite{hinze1995projection} Section 6.3.}.

\begin{figure}
\noindent

\begin{align*}
[C_{1} \alpha_{1} | \dots | C_{n} \alpha_{n}] \pmeet& [C_{1} \gamma_{1} | \dots | C_{n} \gamma_{n}] \\
    & \quad = [C_{1} (\alpha_{1} \pmeet \gamma_{1}) | \dots | C_{n} (\alpha_{n} \pmeet \gamma_{n})] \\
[C_{1} \alpha_{1} | \dots | C_{n} \alpha_{n}] \sqcup& [C_{1} \gamma_{1} | \dots | C_{n} \gamma_{n}] \\
    & \quad = [C_{1} (\alpha_{1} \sqcup \gamma_{1}) | \dots | C_{n} (\alpha_{n} \sqcup \gamma_{n})]
\end{align*}% what

\begin{align*}
\langle \alpha_{1} \otimes \dots \otimes \alpha_{n} \rangle \pmeet& \langle \gamma_{1} \otimes \dots \otimes \gamma_{n} \rangle \\
    & \quad = norm \big( \langle (\alpha_{1} \pmeet \gamma_{1}) \otimes \dots \otimes (\alpha_{n} \pmeet \gamma_{n})\rangle \big) \\
\langle \alpha_{1} \otimes \dots \otimes \alpha_{n} \rangle \sqcup& \langle \gamma_{1} \otimes \dots \otimes \gamma_{n} \rangle \\
    & \quad = \langle (\alpha_{1} \sqcup \gamma_{1}) \otimes \dots \otimes (\alpha_{n} \sqcup \gamma_{n})\rangle
\end{align*}
\caption[Conjunction and Disjunction for Projections 2]{Conjunction and Disjunction for Projections on Products and Sums}
\label{fig:conjDisSum}
\end{figure}

For conjunction of projections on recursive types we have to perform additional
analysis to ensure that we maintain uniformity, which is the property that the
demand on the `recursive call' of the type is equal to the demand on the type
itself (as mentioned in Section \ref{sec:projSem}). The subtlety is due to the
fact that performing conjunction on two recursive types might result in demands
that differ on the `head' of the value from the demand on the recursive call
\citep{kubiak, hinze1995projection}.

A simple example is when we perform conjunction on the \<H\> and \<T\>
projections on lists (from Figure \ref{fig:listContexts}). If we naively
perform conjunction as
\<(\mu\beta.\alpha) \pmeet (\mu\beta.\gamma)
= \mu\beta.\alpha \pmeet \gamma\>, we
arrive at \<HT\>, while this may seem like the correct result it is actually
unsafe! This is clear when applying these projections to the list \<xs = 1 :
\bot : []\>

\begin{haskell*}
H xs  &\equiv& 1 : \_       &&\hscom{Head strictness forces the first element} \\
T xs  &\equiv& \_ : \_ : [] &&\hscom{Tail strictness forces the spine} \\
HT xs &\equiv& \bot         &&\hscom{\<HT\> forces all elements and the spine}
\end{haskell*}

The reason for the differing results is that \<H\> is not strict in the recursive
part of the list, but \<T\> is, and being head strict is \emph{not} the same
as requiring all elements of a structure, as evidenced by the following small
program

\begin{haskell*}
f xs = a + b \hswhere{%
                a &=& head xs \\
                b &=& length xs
             }
\end{haskell*}

The demand on the input list \<xs\> is the conjunction of the demands for
\<head\> and \<length\> but it is clear to see that \<f\> does not require its
input list to be fully defined, making it unsafe for \<H \pmeet T\> to result
in \<HT\>.

\hfill$\Box$


The simplest way to maintain uniformity is to take the least upper bound
($\sqcup$) of the two projections.  However, this would be too conservative and
we would lose out on some strictness information that is present. Fortunately,
Hinze provides us with a method that allows us to use more accurate
approximations when it is safe to do so, relying on \join only when necessary
(the last guard in Figure \ref{fig:conjRec}). The idea is to use versions of
$\sqcup$ and $\pmeet$ that ignore all demands except those on the recursive
calls; these are written as $\sqcup'$ and $\pmeet'$. We then see where in the
lattice the result (\<conj\>) resides, performing the corresponding
approximation, defaulting to the always safe \<\mu\beta.\alpha \sqcup
\gamma\>. For details on how this method was derived and a proof of its
safety, see \cite{hinze1995projection} Section 6.4.

\begin{figure}
\begin{haskell*}
(\mu\beta.\alpha) \sqcup& (\mu\beta.\gamma) &&= \quad \mu\beta.\alpha \sqcup \gamma &&\ \\
(\mu\beta.\alpha) \pmeet& (\mu\beta.\gamma) &&| \quad conj \sqsubseteq \beta_{1} \pmeet' \beta_{2} &&= norm \big(\mu\beta.\alpha \pmeet \gamma \big) \\
\quad                   &   \               &&| \quad conj \sqsubseteq \beta_{1} \sqcup' \beta_{1} \pmeet' \beta_{2} &&= \mu\beta.\alpha \sqcup \alpha \pmeet \gamma \\
\quad                   &   \               &&| \quad conj \sqsubseteq \beta_{1} \pmeet' \beta_{2} \sqcup' \beta_{2} &&= \mu\beta.\alpha \pmeet \gamma \sqcup \gamma \\
\quad                   &   \               &&| \quad conj \sqsubseteq \beta_{1} \sqcup' \beta_{2} &&= \mu\beta.\alpha \sqcup \gamma
\hswhere{%
    conj &= (norm(\alpha[\beta \mapsto \beta_{1}])) \pmeet' (norm(\gamma[\beta \mapsto \beta_{2}])) \\
    }
\end{haskell*}
\caption[Conjunction and Disjunction for Projections 3]{Conjunction and Disjuntion for Projections on Recursive Types}
\label{fig:conjRec}
\end{figure}



\section{Review of the worker/wrapper transformation}
\label{sec:worker/wrapper}

\section{Deriving Strategies}
\label{sec:deriveStrat}

\section{Future Work}

\section{Conclusions}



\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}


\end{document}
